---
title: "Lab 05 - Wrangling spatial data"
author: "Enhui Wang"
date: "2025.02.10"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1
There are 3 Denny's location and 2 La Quinta's location in Alaska.
```{r}
#filter the Denny’s data frame for Alaska (AK)
dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)

#filter the La Quinta’s data frame for Alaska (AK)
lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)
```


### Exercise 2
There are 6 pairings.
```{r}
#the number of pairings
num_pairings <- nrow(dn_ak) * nrow(lq_ak)
num_pairings
```

### Exercise 3
There are 6 pairings.
```{r}
#join the data on Denny’s and La Quinta locations in Alaska
dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
dn_lq_ak
```

### Exercise 4

There are 6 observations.
```{r}
nrow((dn_lq_ak))
```
### Exercise 5
We use mutate() function. 
```{r}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}
```
...

### Exercise 6
```{r}
#generate all combinations of Denny's and La Quinta locations.
dn_lq_ak <- expand.grid(
  dn_id = 1:nrow(dn_ak),# Denny's locations
  lq_id = 1:nrow(lq_ak) # La Quinta locations
)
#get the coordinates for each  pair of Denny’s and La Quinta locations
dn_lq_ak <- dn_lq_ak %>%
  mutate(
    dn_lat = dn_ak$latitude[dn_id],
    dn_lon = dn_ak$longitude[dn_id],
    lq_lat = lq_ak$latitude[lq_id],
    lq_lon = lq_ak$longitude[lq_id]
  )
#use the haversine function to calculate the distance
dn_lq_ak <- dn_lq_ak %>%
  mutate(
    distance = haversine(dn_lon, dn_lat, lq_lon, lq_lat)
  )
# check the updated data frame
head(dn_lq_ak)
```
### Exercise 7
The minimum distance is 2.035km.
```{r}
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(dn_id) %>% #groups the data by each unique Denny's location
  summarize(closest = min(distance))#calculates the minimum distance between a Denny's and La Quinta for each Denny’s location.
print(dn_lq_ak_mindist )
```
### Exercise 8
```{r}
#summary statistics
summary_stats <- summary(dn_lq_ak_mindist$closest)
print(summary_stats)

#visualization histogram with density
library(ggplot2)

ggplot(dn_lq_ak_mindist, aes(x = closest)) +
  #create a histogram
  geom_histogram(aes(y = ..density..), bins = 10, fill = "purple", alpha = 0.7, color = "black") +
  #create a density curve
  geom_density(color = "yellow") +
  labs(
    title = "Distribution of the distances Denny’s and the nearest La Quinta locations in Alaska",
    x = "Minimum Distance (km)",
    y = "Density"
  )
```

### Exercise 9
The closest Denny’s and La Quinta pairing is 1.779 km. The distribution is positively skewed. 
```{r}
#filter Denny’s and La Quinta Data Frames for NC
dn_nc <- dennys %>%
  filter(state == "NC")
# filter La Quinta for North Carolina
lq_nc <- laquinta %>%
  filter(state == "NC") 

#join data frames to get a complete list of all possible pairings
dn_lq_nc <- expand.grid(
  dn_id = 1:nrow(dn_nc),# Denny's locations
  lq_id = 1:nrow(lq_nc) # La Quinta locations
) %>%
  mutate(
    dn_lat = dn_nc$latitude[dn_id],
    dn_lon = dn_nc$longitude[dn_id],
    lq_lat = lq_nc$latitude[lq_id],
    lq_lon = lq_nc$longitude[lq_id]
  )
#calculate the distances between all possible pairings of Denny’s and La Quinta in NC
dn_lq_nc <- dn_lq_nc %>%
  mutate(
    distance = haversine(dn_lon, dn_lat, lq_lon, lq_lat)
  )

#find the minimum distance between each Denny’s and La Quinta location
dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(dn_id) %>%
  summarize(closest = min(distance))

#summary statistics
summary_stats <- summary(dn_lq_nc_mindist$closest)
print(summary_stats)

#visualization histogram with density 
ggplot(dn_lq_nc_mindist, aes(x = closest)) +
  #create a histogram
  geom_histogram(aes(y = ..density..), bins = 10, fill = "purple", alpha = 0.7, color = "black") +
  geom_density(color = "yellow") +
  labs(
    title = "Distribution of the distances Denny’s and the nearest La Quinta locations in North Carolina",
    x = "Minimum Distance (km)",
    y = "Density"
  ) 
```

### Exercise 10
The closest Denny’s and La Quinta pairing is 0.016 km. The distribution is positively skewed, and the most distances are concentrated near the lower range. Overall, Denny’s and La Quinta locations in Texas are located very close to each other.
```{r}
#filter Denny’s and La Quinta Data Frames for Texas
dn_tx <- dennys %>%
  filter(state == "TX")
# filter La Quinta for Texas
lq_tx<- laquinta %>%
  filter(state == "TX") 

#join data frames to get a complete list of all possible pairings
dn_lq_tx <- expand.grid(
  dn_id = 1:nrow(dn_tx), # Denny's locations
  lq_id = 1:nrow(lq_tx)  # La Quinta locations
) %>%
  mutate(
    dn_lat = dn_tx$latitude[dn_id],
    dn_lon = dn_tx$longitude[dn_id],
    lq_lat = lq_tx$latitude[lq_id],
    lq_lon = lq_tx$longitude[lq_id]
  )
#calculate the distances between all possible pairings of Denny’s and La Quinta in Texas
dn_lq_tx <- dn_lq_tx %>%
  mutate(
    distance = haversine(dn_lon, dn_lat, lq_lon, lq_lat)
  )

#find the minimum distance between each Denny’s and La Quinta location
dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(dn_id) %>%
  summarize(closest = min(distance))

#summary statistics
summary_stats <- summary(dn_lq_tx_mindist$closest)
print(summary_stats)

#visualization histogram with density 
ggplot(dn_lq_tx_mindist, aes(x = closest)) +
  #create a histogram
  geom_histogram(aes(y = ..density..), bins = 10, fill = "purple", alpha = 0.7, color = "black") +
  geom_density(color = "yellow") +
  labs(
    title = "Distribution of the distances Denny’s and the nearest La Quinta locations in Texas",
    x = "Minimum Distance (km)",
    y = "Density"
  ) 
```

### Exercise 12
I choose Arizona for this exercise. The closest Denny’s and La Quinta pairing is 0.021 km. The distribution is positively skewed, and the most distances are concentrated near the lower range. Overall, Denny’s and La Quinta locations in Arizona are located very close to each other.

```{r}
#filter Denny’s and La Quinta Data Frames for Arizona
dn_az <- dennys %>%
  filter(state == "AZ")
# filter La Quinta for Arizona
lq_az<- laquinta %>%
  filter(state == "AZ") 

#join data frames to get a complete list of all possible pairings
dn_lq_az <- expand.grid(
  dn_id = 1:nrow(dn_az), # Denny's locations
  lq_id = 1:nrow(lq_az)  # La Quinta locations
) %>%
  mutate(
    dn_lat = dn_az$latitude[dn_id],
    dn_lon = dn_az$longitude[dn_id],
    lq_lat = lq_az$latitude[lq_id],
    lq_lon = lq_az$longitude[lq_id]
  )
#calculate the distances between all possible pairings of Denny’s and La Quinta in Arizona
dn_lq_az <- dn_lq_az %>%
  mutate(
    distance = haversine(dn_lon, dn_lat, lq_lon, lq_lat)
  )

#find the minimum distance between each Denny’s and La Quinta location
dn_lq_az_mindist <- dn_lq_az %>%
  group_by(dn_id) %>%
  summarize(closest = min(distance))

#summary statistics
summary_stats <- summary(dn_lq_az_mindist$closest)
print(summary_stats)

#visualization histogram with density 
ggplot(dn_lq_az_mindist, aes(x = closest)) +
  #create a histogram
  geom_histogram(aes(y = ..density..), bins = 10, fill = "purple", alpha = 0.7, color = "black") +
  geom_density(color = "yellow") +
  labs(
    title = "Distribution of the distances Denny’s and the nearest La Quinta locations in Arizona",
    x = "Minimum Distance (km)",
    y = "Density"
  ) 
```

### Exercise 13

Mitch Hedberg’s joke most likely to hold true in Texas. Because the minimum distance is 0.016 km, suggesting the locations overlap with each other. According to the visualization, the distribution is relatively highly positively skewed in Texas compared to other states, suggesting the majority of Denny's and La Quinta locations are close to each other. 
